---
title: "Trinity and Transrate Tutorial"
author: "Nate Olson"
date: "May 2, 2016"
output:
  html_document:
    keep_md: true
bibliography: tutorial.bib 
---
```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(knitr)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(GGally)
```

```{r source_dir, warning=FALSE, message=FALSE, echo = FALSE}
results_list <- list(
      no_mods = "data/no_mods/Trinity_sequences.Trinity.fixed/",
      corrected = "data/corrected/Assembly4_Corrected.Trinity.fixed/",
      trimmed = "data/trimmed/Assembly2_Trimmed.Trinity.fixed/",
      trimmedcorrected = "data/trimmedcorrected/Assembly3_TrimmedCorrected.Trinity.fixed/"
)
```

## Background 
* Transcriptome
* Transcriptome assembly
      * Trinity [@haas2013novo]
* Transcriptome assembly evaluation
      * Reference based
      * Read based
      * Transrate - contig level quality score [@smith2015transrate]
* C. elegans transcriptome assembly and evaluation (what we did)

## Assembly
### Source data  
Sequence data identified by searching SRA for _C. elegans_ and limiting results to RNA, identified a study for an Illumina HiSeq 2000, 94 bp paired end run. Accession SRR2969230 was used to generate and evaluate the assembly.

> 
Caenorhabditis elegans Transcriptome or Gene expression BioProject
Abstract: Insulin/IGF-1 signaling (IIS) is a critical regulator of the most important biological decisions, from rates of growth, development, and metabolism, to reproduction and longevity. However, the neuron-specific IIS/FOXO targets that regulate these functions are still unknown. By isolating C. elegans neurons for transcriptional profiling, we identified both the wild-type and IIS/FOXO neuronal transcriptomes. Here we deposit the sequences for wt adult neurons and IIS mutants, as well as for isolated adult mechanosensory neurons.

Relevant _C. elegans_ Reference Sequences  

* Transcriptome: ftp://ftp.ensemblgenomes.org/pub/release-31/metazoa/gtf/caenorhabditis_elegans/Caenorhabditis_elegans.WBcel235.31.gtf.gz   
* Genome: ftp://ftp.ensemblgenomes.org/pub/release-31/metazoa/fasta/caenorhabditis_elegans/dna/Caenorhabditis_elegans.WBcel235.31.dna.chromosome.*.fa.gz   
* Proteome: ftp://ftp.ensemblgenomes.org/pub/release-31/metazoa/fasta/caenorhabditis_elegans/pep/Caenorhabditis_elegans.WBcel235.31.pep.all.fa.gz    

### How to run trinity  
* Link to website  
* Parameter description
* Command used to generate _C. elegans_ assembly

### Assembly results  
Assembly metrics calculated using Transrate, see http://hibberdlab.com/transrate/metrics.html for a description of the different metrics.  

```{r}
assembly_metrics_df <- list(no_mods = "data/no_mods/", 
                            corrected = "data/corrected/",
                            trimmed = "data/trimmed/", 
                            trimmedcorrected = "data/trimmedcorrected/") %>% 
      map(paste0, "assemblies.csv") %>%  
      map_df(read_csv, .id = "read_set") %>% 
      select(-assembly) %>% 
      gather("metric","value", -read_set) 
```

__TODO__ Need to workout `read_set` order
```{r}
assembly_metrics_df %>% 
      filter(metric %in% c("n_seqs", "smallest", "largest", 
                           "n50", "n_with_orf")) %>%
      spread(metric, value) %>% 
      select(read_set, n_seqs, smallest, n50, largest, n_with_orf) %>% 
      kable(digits = 2,
            caption = "Trinity _C. elegans_ transcriptome assembly summary metrics.")
```

## Assembly evaluation
### Running transrate
* Installation
      * Precompiled binaries available from website (http://hibberdlab.com/transrate/), along with instructions to build from source
      * also command to install dependencies `transrate --install-deps type` where type can be either `all`, `read`, or `ref`.
      * See Transrate website for additional information for install. 


__Command Line__  

_Read Based_

```
transrate \
  --left=SRR2969230_1.fastq \
  --right=SRR2969230_2.fastq \
  --assembly=Trinity_sequences.Trinity.fixed.fasta \
  --output=transrate/
```

_Reference Based_
```
transrate \
  --reference=Caenorhabditis_elegans.WBcel235.31.pep.all.fa \
  --assembly=Trinity_sequences.Trinity.fixed.fasta \
  --output=transrate/
```

### Transrate results
The unmodified read set was passed as input to Transrate, 
potentally biasing some metric results towards the unmodifed assembly.

#### run time
__TODO__ Add runtime for other assemblies and reference based evaluations  

- real    890m58.016s
- user    6665m50.869s
- sys     11m5.890s

#### Assembly Scores
Read trimming and error correction resulted in the highest weighted assembly score, 
while having the lowest contig score cutoff, 
but a lower optimal score.  

```{r}
assembly_metrics_df %>% 
      filter(metric %in% c("score", "optimal_score", "cutoff", "weighted")) %>%
      spread(metric, value) %>% 
      kable(digits = 2,
            caption = "Trinity _C. elegans_ transcriptome Transrate score summary.")
```


#### Assembly Score Optimization
```{r}
assembly_score_opt <- results_list %>% map(paste0,"assembly_score_optimisation.csv") %>% 
      map_df(read_csv, .id = "read_set")
```

Relationship between the cutoff for contig score and assembly score.
```{r}
assembly_score_opt %>% 
      ggplot() + geom_path(aes(x = cutoff, y = assembly_score, color = read_set)) +
            theme_bw() +
            labs(x = "Contig Score Threshold", y = "Assembly Score")
```

### Contigs
```{r}
contig_stat <- results_list %>% map(paste0,"contigs.csv") %>% 
      map_df(read_csv, .id = "read_set")

contig_cutoff <- assembly_metrics_df %>% filter(metric == "cutoff") %>% 
      select(-metric) %>% rename(cutoff = value)
contig_stat <- contig_stat %>% left_join(contig_cutoff) %>% 
      mutate(contig_filt = cutoff < score)
```


__Contig Score Distribution__
Trimming reads resulted in a higher proportion of contigs with scores greater than 0.25 then assemblies using unmodified reads or only error corrected reads.  

```{r}
contig_stat %>% 
      ggplot() + geom_density(aes(x = score, color = read_set, fill = read_set), 
                              alpha = 0.25) + theme_bw()
```


Relationship between contig score and length. 
```{r score_length, fig.cap = "2D histogram of contig score and legnth, with color indicating abundance. Grey dotted lines are the contig score cutoff for optimal assembly score."}
contig_stat %>% 
      ggplot() + geom_hex(aes(x = length, y = score)) +
            geom_hline(aes(yintercept = cutoff), 
                       color = "grey60", linetype = 2) +
            theme_bw() +
            labs(x = "Contig Length (bp)", y = "Contig Score") +
            facet_wrap(~read_set)
```

Proportion of bases in reads map to the transcript that agree with the transcript contig. 

```{r p_good_density, fig.cap = "Histogram of the proportion of bases in agreement with contig and contig length."}
contig_stat %>% 
      ggplot() + geom_histogram(aes(x = p_good)) +
            theme_bw() +
            labs(x = "Proportion of Bases Aggree with Transcript",
                 y = "Count") +
            facet_grid(contig_filt~read_set) +
            theme(legend.position = "bottom")
```

Proportion of bases coverage by mapped reads.
```{r}
contig_stat %>% 
      ggplot() + geom_histogram(aes(x = p_bases_covered)) +
            theme_bw() +
            labs(x = "Proportion of Covered Bases", 
                 y = "Count") +
            facet_grid(contig_filt~read_set) +
            theme(legend.position = "bottom")
```

```{r}
contig_stat %>% 
      ggplot() + geom_histogram(aes(x = p_seq_true)) +
            theme_bw() +
            labs(x = "Proportion Seq Correct", 
                 y = "Count") +
            facet_grid(contig_filt~read_set) +
            theme(legend.position = "bottom")
```

```{r}
contig_stat %>% 
      ggplot() + geom_histogram(aes(x = p_not_segmented)) +
            theme_bw() +
            labs(x = "Proportion of Non-Segmented Contigs", 
                 y = "Count") +
            facet_grid(contig_filt~read_set) +
            theme(legend.position = "bottom")
```

The individual parameters used to calculate the contig score are weakly correlated. 
Correlation values are similar to thoes presented in the Transrate publication. 

```{r}
for(i in names(results_list)){
      pl <- contig_stat %>% filter(read_set == i) %>% 
            select(p_good, p_bases_covered, p_seq_true, p_not_segmented) %>% 
            ggcorr(method = c("pairwise","spearman"),label = TRUE) + 
                  ggtitle(i)
      print(pl)
}
```


```{r}
minmap <- read_tsv("../analysis/celegans/minimap_trinity_cds.pfa",
                   col_names = c("ref_id","ref_tran_len","ref_start","ref_end",
                                 "break","t_id","t_contig_len","t_start","t_end",
                                 "ref_aln_len", "t_aln_len","aln_score","aln_met"))
```

Total transcripts in reference with hits `r unique(minmap$ref_id) %>% length()`

Number of contigs mapping to each contig
```{r}
count_sum <- minmap %>% group_by(ref_id, t_id) %>% summarize(aln_count = n()) 
count_sum %>% filter(aln_count < 10) %>% 
      ggplot() + geom_density(aes(x = aln_count))
```

Contigs map to most contigs more than once with a number of contigs having more than 10 and with over 300 aligning to one reference transcript.

```{r}
count_sum %>% filter(aln_count > 10) %>% .$ref_id %>% unique() %>% length()
```

Looking at the contig with 300
```{r}
#count_sum 
```

```{r}
contig_stat_map <- minmap %>% mutate(contig_name = str_replace(t_id, "|","_")) %>% left_join(contig_stat)
contig_stat_map <- contig_stat_map %>% mutate(aln_ratio = ref_aln_len/t_aln_len)
```

```{r}
# ggplot(contig_stat_map) + geom_point(aes(x = aln_ratio, y = score))
```

__TODO__
* Other factors of interest
* Expore seq information for example high scoring and low scoring contigs

### Transrate-paper method results
* Github repository for paper 
* ability to rerun analysis - reproducible research
* https://github.com/blahah/transrate-paper

## Conclusions
* How did Trinity do?
* How did transrate compare to other methods?
* Recommendations for transcriptome assembly and evaluation

## Session Information
__Session Info__
```{r s_info}
s_info <- devtools::session_info()
print(s_info$platform)
kable(s_info$packages)
```
## References
